### 2015-10-16 manuscript.Rnw
various edits

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \citep{Pennell2013}. The development of these methods has
contributed to the need for a greater availability of high-quality phylogenetic
information across life. However, phylogenetic information is scattered, often
only available as image files within publications, and the variety of file
formats used to store phylogenetic data makes it difficult for researchers to
access, synthesize, and integrate this information into their own research
(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
practices).

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by OTL. However, resources and methods are being developed to add
branch lengths information to these topological trees
\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
phylogenetically equivalent species to increase overlap between chronograms and
species trait data \citep{Pennell2015}. Without branch lengths, these trees
remain useful to illustrate relationships among species, or to map a trait on a
phylogeny for instance.

@@

To get the relationships among a set of taxa from the Open Tree, the taxa first
need to be matched against the Open Taxonomy using the TNRS. This step retrieves
the taxa identifiers that will be used to extract the relationships for the set
of requested taxa.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and identify the trees that
contribute to the synthetic tree. These trees provide a useful resource to
reproduce or expand on an already published analysis.

@@

<<get_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
                                    value = "felidae")
cat_studies
@

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. As many methods for comparative analyses are implemented,
and because it is a relatively easy to use programming language, R greatly
facilitates the analysis of large datasets, and allows researchers to combine
methods in novel ways. Additionally, as more data is made available online and
accessible using APIs, several R packages have been developed to take advantage of
these datasets, especially spearheaded by the organization rOpenSci
\url{https://ropensci.org}. \texttt{rotl} contributes to these initiatives and
data from OTL can be combined with other sources easily. For instance in
Figure~\ref{fig:cat_map}, we show how we can obtain a map of the occurrences for
some of the cat species (genus \textit{Lynx}) included in the phylogeny above
using records for these species found in GBIF (code is available in
supplementary material).

<<cat_map, fig.height=4.2, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
## load required packages
suppressPackageStartupMessages(library(rgbif))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(maps))

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
\url{https://cran.r-project.org/package=rotl}, and development
versions are available from GitHub
\url{https://github.com/ropensci/rotl}.

>>>>

Generating phylogenies has been greatly facilitated by advances in sequencing
and computing technologies. In parallel, new phylogenetic comparative methods
have been developed and make use of these phylogenies to explore fundamental
questions about the origin of biodiversity including the evolution of
morphological and ecological traits, the spatio-temporal variation in speciation
rates, or both \citep{OMeara2012,Pennell2013}. The development of these methods
has contributed to the need for a greater availability of high-quality
phylogenetic information across life. However, phylogenetic information is
scattered, often only available as image files within publications, and the lack
of standardization to store and represent phylogenetic data makes it difficult
for researchers to access, synthesize, and integrate this information into their
own research (\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for
suggestions of best practices).

@@

The synthetic tree currently does not have any branch lengths associated with
it, therefore parametric comparative methods cannot be used directly on the
trees returned by OTL. However, resources and methods are being developed to add
branch lengths to these topological trees \citep[e.g.,][]{Ksepka2015} or use
topological trees to identify phylogenetically equivalent species to increase
overlap between chronograms and species trait data \citep{Pennell2015}. Without
branch lengths, these trees are nonetheless useful to illustrate relationships
among species, or to map a trait on a phylogeny for instance.

@@

To get the relationships among a set of taxa from the Open Tree, the taxa first
need to be matched against the Open Taxonomy using the TNRS. This step retrieves
the identifiers that will be used to extract the relationships for the set of
requested taxa.

@@

\texttt{rotl} can also be used to retrieve trees accompanying studies that have
been submitted through the curator interface, and identify the trees that
contribute to the synthetic tree. These trees provide a useful resource to
reproduce or expand on an already published analysis, or to explore how the
elucidation of relationships within a clade has changed through time.

@@

<<get_cats, purl=FALSE, fig.cap="Tree from \\citealt{Johnson2006} obtained from OTL">>=
cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
                                    value = "felidae")
cat_studies
@

@@

In recent years, R has become part of the toolbox of many researchers in
evolutionary biology. R greatly facilitates the analysis of large datasets, and
allows researchers to combine methods in novel ways because many methods for
comparative analyses are implemented, and because it is a relatively easy to use
programming language. Additionally, as more data is made available online and
accessible using APIs, several packages have been developed to interact and
download these datasets directly in R enabling direct and reproducible analyses.
Notably, the organization rOpenSci \url{https://ropensci.org} has fostered a
community of researchers who develop tools and methods to facilitate the use of
open data and broaden the adoption of open science practices in
general. \texttt{rotl} contributes to this initiative, and data from OTL can be
combined with other sources easily. For instance in Figure~\ref{fig:cat_map}, we
show how we can obtain a map of the occurrences for some of the cat species
(genus \textit{Lynx}) included in the phylogeny above using records for these
species found in GBIF (code available in Appendix). Another example is available
in the package vignette ``Data mashups''.

<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
## load required packages
suppressPackageStartupMessages(library(rgbif))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(maps))

@@

\texttt{rotl} is free, open source, and released under a Simplified
BSD license. Stable versions are available from the CRAN repository
\url{https://cran.r-project.org/package=rotl}, and development
versions are available from GitHub
\url{https://github.com/ropensci/rotl}. The source for this manuscript is
available at \url{https://github.com/fmichonneau/rotl-ms}.
